---
title: "Analyses for James et al. 2022: An in-depth statistical analysis of the COVID- 19 pandemic's initial spread in the WHO African region"
date: "06/01/2022"
author: 'Ananthu James, Jyoti Dalal, Jessie Abbate'
output: 
  html_document: 
  #fig_height: 10
  #fig_width: 12.5
  editor_options: 
  chunk_output_type: console
---
  
```{r, include=FALSE}
rm(list = ls()) 
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)
```

```{r, include=FALSE}
require(pacman)
library(pacman)
p_load(char = c("devtools", "jtools", "ggstance", "grid", "gridExtra", 
                "corrplot", # for correlation plots
                "FactoMineR", "factoextra", # for PCA analysis
                "kableExtra", # for nice html tables
                "GGally", "RColorBrewer", 
                "glmulti","linelist", "MASS",
                "tidyverse", 'readr')) # for everything
devtools::install_github("raivokolde/pheatmap") #for adaptable heatmaps, developer version
library(pheatmap)
```


```{r}

df <- read.csv("./46country_whoafro_datafile.csv", 
               header = TRUE, sep = ",", dec = ".") 

df = df[-(which(df$country=="United Republic of Tanzania")), ]  # removing due to poor data quality

df$max_weekly_attackrate <- df$max_weekly_cases/(df$pop_thous/1000)
df$max_monthly_attackrate <- df$max_monthly_cases/(df$pop_thous/1000)

df$CFR<-df$deaths/df$cases

df$days_to_50cases<-as.numeric((as.Date(df$day_50th_case)-as.Date(df$day_first_case)))

```


# 1. Descriptive Statistics on Epidemic Indicators


```{r}
# recode long country names
df$country
df$country[df$country=="Democratic Republic of the Congo"]<-"DRC"
df$country[df$country=="Central African Republic"]<-"CAR"
df$country[df$country=="Congo (Republic of)"]<-"Congo"
df$country[df$country=="Sao Tome and Principe"]<-"STP"
df$country[df$country=="Equatorial Guinea"]<-"Eq Guinea"

dfhm<-as.matrix(df[,c("cases_per_million", "max_monthly_attackrate", "CFR", "start_delay", "days_to_50cases")])

row.names(dfhm)<-c(df$country)
colnames(dfhm)<-c("Cumulative Attack Rate", "Maximum Attack Rate","CFR","Start Delay","Growth Period")

pheatmap(t(dfhm),scale="row", cluster_rows=FALSE, cluster_cols=FALSE,legend=TRUE,fontsize=7,cellwidth = 8,cellheight = 8, color = colorRampPalette(c("blue", "white", "red"))(30))

```


- Countries for which values for variables 'Stringency_Mean_Min', 'Stringency_Max_Mean' are missing:

```{r}
#df<-droplevels(df[which(df$deaths>=10),]) # Removing countries with less than 10 deaths

df$country[which(is.na(df$Stringency_Mean_Min))]
df$country[which(is.na(df$Stringency_Max_Mean))]

```


- Countries for which values for variables 'tourism_arrivals' and 'tourism_arrivals_standardized' values are missing

```{r}
df$country[which(is.na(df$Tourism_Arrivals))]

```

-Countries with fewer than 10 deaths recorded

```{r}
df$country[which(df$deaths<10)]
```

# 2. Exploratory Analysis (46 countries)
 
## Explanatory variables  

```{r}
df1 <- as.data.frame(df$country)

df1$Tourism_arrivals_percap = df$Tourism_Arrivals_standardised_by_population
df1$Tourism_dollars_percap = df$Tourism_receipts_standardised_by_population
df1$Stringency_day1 = df$Strigency_Index_First_case_day
df1$Stringency_mean = df$Stringency_Mean_Min
df1$neighbors_attack_rate = df$cases_per_million_population_in_neighbors
df1$GDP_percap = df$GDP_per_capita
df1$Fishing_volume_percap = df$Capture_fisheries_standardised_by_population
df1$Preparedness = df$Prepardness_Index_Total_Average
df1$Latitude = df$latitude
df1$Urbanization = as.numeric(df$Urban_population_percentage)
df1$Children_proportion = df$population_age_between_0_14_years_percentage
df1$Male_proportion = df$male_pop_frac
df1$Pop_density = as.numeric(df$Pop_Density_per_sq_km) 

dfhm_explan<-as.matrix(df1[,2:dim(df1)[2]]) 
row.names(dfhm_explan)<-c(df1$country)
colnames(dfhm_explan)<-c("Tourism Arrivals","Tourism Revenue","Initial Stringency","Mean Stringency","Neighbor Attack Rate", "Per capita GDP","Per capita Fishing","Preparedness","Latitude","Urbanization","Proportion Under 15","Proportion Male","Population Density")

pheatmap(t(dfhm_explan),scale="row", cluster_rows=FALSE, cluster_cols=FALSE,legend=TRUE,fontsize=7,cellwidth = 8,cellheight = 8, color = colorRampPalette(c("blue", "white", "red"))(30),na_col="grey")
```

## Format data for analysis

```{r}

dfclip = df1 %>% 
 mutate(across(where(is.integer), as.numeric)) 

names(dfclip)[which(purrr::map_df(dfclip, class)!="numeric")]     
dfclip[,which(purrr::map_df(dfclip, class)=="numeric")]

dfclip<-dfclip[,which(purrr::map_df(dfclip, class)=="numeric")]
which(is.na(dfclip))

```

## Data Imputation

```{r, include=TRUE}

df2<-as.data.frame(dfclip)

# impute Tourism_arrivals_percap
hist(log10(dfclip$Tourism_arrivals_percap))
df2[which(is.na(dfclip$Tourism_arrivals_percap)),"Tourism_arrivals_percap"]<- median(dfclip$Tourism_arrivals_percap,na.rm=TRUE)

# impute Tourism_dollars_percap
hist(log10(dfclip$Tourism_dollars_percap))
df2[which(is.na(dfclip$Tourism_dollars_percap)),"Tourism_dollars_percap"]<- median(dfclip$Tourism_dollars_percap,na.rm=TRUE)

# impute Stringency_day1
hist(sqrt(dfclip$Stringency_day1))
df2[which(is.na(dfclip$Stringency_day1)),"Stringency_day1"]<-(median(sqrt(dfclip$Stringency_day1),na.rm=TRUE))^2 # I don't expect this to be any different from imputing just the median 

# impute Stringency_mean
hist((dfclip$Stringency_mean))
df2[which(is.na(dfclip$Stringency_mean)),"Stringency_mean"]<-median(dfclip$Stringency_mean,na.rm=TRUE)
```

## Transform heavily skewed explanatory variables for (pairwise) correlation  

```{r} 

hist(log10(df2[,"Tourism_arrivals_percap"]))
hist(log10(df2$Tourism_dollars_percap))
hist(sqrt(df2$Stringency_day1))
hist((df2$Stringency_mean))
hist(log10(df2$neighbors_attack_rate))
hist(log10(df2$GDP_percap))
hist(log10(df2$Fishing_volume_percap))
hist((df2$Preparedness))
hist((df2$Latitude))
hist((df2$Urbanization))
hist((df2$Children_proportion))
hist((df2$Male_proportion))
hist(log10(df2$Pop_density))

df2$Tourism_arrivals_percap_tr<-(log10(df2$Tourism_arrivals_percap))
df2$Tourism_dollars_percap_tr<-(log10(df2$Tourism_dollars_percap))
df2$Stringency_day1_tr<-(sqrt(df2$Stringency_day1))
df2$neighbors_attack_rate_tr<-(log10(df2$neighbors_attack_rate+1))  # 1 is added to allow log transformation including the 6 island nations (neighbors = 0). 
df2$GDP_percap_tr<-(log10(df2$GDP_percap))
df2$Fishing_volume_percap_tr<-(log10(df2$Fishing_volume_percap))
df2$Pop_density_tr<-(log10(df2$Pop_density))
```

## Correlation matrix among explanatory variables

```{r}
names(df2)

df2_tr<- df2 %>% select(Tourism_arrivals_percap_tr,Tourism_dollars_percap_tr,Stringency_day1_tr,Stringency_mean,neighbors_attack_rate_tr,GDP_percap_tr, Fishing_volume_percap_tr,Preparedness,Latitude,Urbanization,Children_proportion,Male_proportion,Pop_density_tr)

names(df2_tr)<-c("Tourism Arrivals","Tourism Revenue","Initial Stringency","Mean Stringency","Neighbor Attack Rate", "Per capita GDP","Per capita Fishing","Preparedness","Latitude","Urbanization","Proportion Under 15","Proportion Male","Population Density")

corr_matrix <- round(cor(df2_tr, method = "pearson"), digit = 2) 

corrplot.mixed(corr_matrix, order = "hclust", lower.col = "black", tl.pos = "lt", tl.srt = 45, tl.col = "black", tl.cex = 1, number.cex = 0.5)

```


## PCA (using the non-transformed variables) 

```{r}
df2_not_tr<- df2 %>% select(Tourism_arrivals_percap,Tourism_dollars_percap,Stringency_day1,Stringency_mean,neighbors_attack_rate,GDP_percap, Fishing_volume_percap,Preparedness,Latitude,Urbanization,Children_proportion,Male_proportion,Pop_density)

names(df2_not_tr)<-c("Tourism Arrivals","Tourism Revenue","Initial Stringency","Mean Stringency","Neighbor Attack Rate", "Per capita GDP","Per capita Fishing","Preparedness","Latitude","Urbanization","Proportion Under 15","Proportion Male","Population Density")

res_pca <- PCA(df2_not_tr, ncp = 4, scale.unit = TRUE, graph = FALSE)
eigen_values <- round(res_pca$eig, 4)

knitr::kable(eigen_values, align = "c", 
             caption = "Eigenvalues and explained variance in each Principal Component") 

fviz_eig(res_pca, ncp = 14, choice = "eigenvalue", addlabels = TRUE, ylim = c(0, 5)) +
  geom_hline(yintercept = 1, color = "red", lty = 2) +
  theme_bw()

scree<-fviz_screeplot(res_pca,addlabels=TRUE,ncp=length(df2_not_tr))
scree<-scree + labs(title = NULL,
         x = "PCA Dimensions", y = "% variance explained")+geom_hline(yintercept=(100/length(df2_not_tr)),linetype="dashed",col="red",cex=1.25)+
theme(axis.text=element_text(size=12,face="bold"),
        axis.title=element_text(size=14,face="bold"))
scree
```

-Variable contributions to each PCA Dimension:

```{r}
tab_contrib <- round(res_pca$var$contrib[,1:4], 3)

kable(tab_contrib, align = "c", 
      caption = "Contributions of each variable (%)")

tab_load <- round(sweep(res_pca$var$coord, 2, 
                        sqrt(res_pca$eig[1:ncol(res_pca$var$coord),1]), 
                        FUN = "/"), 3)[,1:4]

kable(tab_load, align = "c", 
      caption = "Loadings of each variable")

fviz_contrib(res_pca, choice = "var", axes = 1)
fviz_contrib(res_pca, choice = "var", axes = 2)
fviz_contrib(res_pca, choice = "var", axes = 3)
fviz_contrib(res_pca, choice = "var", axes = 4)

res.desc <- dimdesc(res_pca, axes = c(1,2,3,4)) 
# Description of dimensions (including p-values)
res.desc$Dim.1
res.desc$Dim.2
res.desc$Dim.3
res.desc$Dim.4

varx <- get_pca_var(res_pca)
corrplot(varx$contrib,is.corr=FALSE,cl.cex=0.5)
corrplot((round(varx$contrib,digits=2)),is.corr=FALSE,cl.cex=0.5, method = "number", sig.level = 0.05, insig = c("blank"), tl.cex = 1.25, number.cex = 0.8,tl.col="black")
corrplot(varx$coord,is.corr=FALSE,cl.cex=0.5)
corrplot((round(varx$coord,digits=2)),is.corr=FALSE,cl.cex=0.5, method = "number", sig.level = 0.05, insig = c("blank"), tl.cex = 1.25, tl.col="black", number.cex = 0.8)
```

### Putting the PC scores and response variables together in a single dataframe for regression and PCA biplot analysis

```{r}

df1$PC1<-res_pca$ind$coord[,1]
df1$PC2<-res_pca$ind$coord[,2]
df1$PC3<-res_pca$ind$coord[,3]
df1$PC4<-res_pca$ind$coord[,4]

names(df1)[1]<-"country"
df_scores<-left_join(df,df1)
territory <- as.factor(df_scores$Territory)
subregion<-as.factor(df_scores$SubRegion)

fviz_pca_biplot(res_pca, axes=c(1,2), invisible=c("var"), habillage = territory, addEllipses = TRUE, label = "var", shape.var = 15) +theme_minimal()

fviz_pca_biplot(res_pca, axes=c(1,2), invisible=c("var"), habillage = subregion, addEllipses = TRUE, label = "var", shape.var = 15) +theme_minimal()

```

# 3. Impact of PCs on Response Variables

## Regression

### CFR

```{r}

mod21<-MASS::glm.nb(deaths~offset(log(cases))+PC1+PC2+PC3+PC4,data=df_scores,init.theta=1.0,link=log)
summary(mod21)   

p_cfr<-summary(mod21)$coefficients[,"Pr(>|z|)"]
p_cfr

mod_21<-stepAIC(mod21)      
summary(mod_21) 

p_best_cfr<-summary(mod_21)$coefficients[,"Pr(>|z|)"]
p_best_cfr

coef(mod_21)
confint(mod_21)
hist(residuals(mod_21))
h21  <- hist(residuals(mod_21))

p21  <- plot_coefs(mod_21, legend.title = "CFR") 

resmod_21 <- residuals(mod_21, type="deviance") 
plot(predict(mod_21), resmod_21) 
plot(log(predict(mod_21)), resmod_21) 
abline(h=0, lty=2)

res_standardised <- rstandard(mod_21)
plot(predict(mod_21), res_standardised) 
plot(log(predict(mod_21)), res_standardised) 
abline(h=0, lty=2)

resP <- residuals(mod_21, type="pearson") 
plot(predict(mod_21), resP) 
plot(log(predict(mod_21)), resP) 
abline(h=0, lty=2)
```


### Attack Rate 

```{r}

mod22<-MASS::glm.nb(cases~offset(log(pop_thousands/1000))+PC1+PC2+PC3+PC4,data=df_scores,init.theta=1.0,link=log)
summary(mod22)

p_attackrate<-summary(mod22)$coefficients[,"Pr(>|z|)"]
p_attackrate

mod_22<-stepAIC(mod22)    
summary(mod_22)  

p_best_attackrate<-summary(mod_22)$coefficients[,"Pr(>|z|)"]
p_best_attackrate

coef(mod_22)
confint(mod_22)
hist(residuals(mod_22))
h22  <- hist(residuals(mod_22))

p22  <- plot_coefs(mod_22, legend.title = "Cumulative attack rate")

resmod_22 <- residuals(mod_22, type="deviance") 
plot(predict(mod_22), resmod_22) 
plot(log(predict(mod_22)), resmod_22) 
abline(h=0, lty=2)

res_standardised <- rstandard(mod_22)
plot(predict(mod_22), res_standardised) 
plot(log(predict(mod_22)), res_standardised) 
abline(h=0, lty=2)

resP <- residuals(mod_22, type="pearson") 
plot(predict(mod_22), resP) 
plot(log(predict(mod_22)), resP) 
abline(h=0, lty=2)

```

### Max Monthly Attack Rate 

```{r}

mod23<-MASS::glm.nb(max_monthly_cases~offset(log(pop_thousands/1000))+PC1+PC2+PC3+PC4,data=df_scores,init.theta=1.0,link=log)
summary(mod23)

p_monthlyattackrate<-summary(mod23)$coefficients[,"Pr(>|z|)"]
p_monthlyattackrate

mod_23<-stepAIC(mod23)
summary(mod_23)

p_best_monthlyattackrate<-summary(mod_23)$coefficients[,"Pr(>|z|)"]
p_best_monthlyattackrate

coef(mod_23)
confint(mod_23)
hist(residuals(mod_23))
h23  <- hist(residuals(mod_23))

p23  <- plot_coefs(mod_23, legend.title = "Max. monthly attack rate")

resmod_23 <- residuals(mod_23, type="deviance") 
plot(predict(mod_23), resmod_23) 
plot(log(predict(mod_23)), resmod_23) 
abline(h=0, lty=2)

res_standardised <- rstandard(mod_23)
plot(predict(mod_23), res_standardised) 
plot(log(predict(mod_23)), res_standardised) 
abline(h=0, lty=2)

resP <- residuals(mod_23, type="pearson") 
plot(predict(mod_23), resP) 
plot(log(predict(mod_23)), resP) 
abline(h=0, lty=2)

```

### Start delay

```{r}

mod1<-MASS::glm.nb(start_delay~PC1+PC2+PC3+PC4,data=df_scores,init.theta=1.0,link=log)

summary(mod1)

p_startdelay<-summary(mod1)$coefficients[,"Pr(>|z|)"]
p_startdelay

mod34<-stepAIC(mod1)
summary(mod34)

p_best_startdelay<-summary(mod34)$coefficients[,"Pr(>|z|)"]
p_best_startdelay

mod34
coef(mod34)
confint(mod34)
hist(residuals(mod34))
h34  <- hist(residuals(mod34))

p34  <- plot_coefs(mod34, legend.title = "Start delay")

resmod34 <- residuals(mod34, type="deviance") 
plot(predict(mod34), resmod34) 
abline(h=0, lty=2)

res_standardised <- rstandard(mod34)
plot(predict(mod34), res_standardised) 
abline(h=0, lty=2)

resP <- residuals(mod34, type="pearson") 
plot(predict(mod34), resP) 
abline(h=0, lty=2)

```


### Initial Growth Period (Days to First 50 Cases)

```{r}

mod1<-MASS::glm.nb(days_to_50cases~PC1+PC2+PC3+PC4,data=df_scores,init.theta=1.0,link=log) # AIC = 402 
summary(mod1)

p_daystofirst50<-summary(mod1)$coefficients[,"Pr(>|z|)"]
p_daystofirst50

mod25<-stepAIC(mod1)
mod25
summary(mod25)

p_best_daystofirst50<-summary(mod25)$coefficients[,"Pr(>|z|)"]
p_best_daystofirst50

coef(mod25)
confint(mod25)
hist(residuals(mod25))
h25  <- hist(residuals(mod25))

p25  <- plot_coefs(mod25, legend.title = "Days to First 50 Cases")

resmod25 <- residuals(mod25, type="deviance") 
plot(predict(mod25), resmod25) 
plot(log(predict(mod25)), resmod25) 
abline(h=0, lty=2)

res_standardised <- rstandard(mod25)
plot(predict(mod25), res_standardised) 
plot(log(predict(mod25)), res_standardised) 
abline(h=0, lty=2)

resP <- residuals(mod25, type="pearson") 
plot(predict(mod25), resP) 
plot(log(predict(mod25)), resP) 
abline(h=0, lty=2)

```

### Visual representation of the regression results 

```{r}

grid.arrange(p21, p22, p23, p34, p25, ncol = 3)

lay = rbind(c(1,2,3),
            c(4,5,NA))

print(grid.arrange(arrangeGrob(p22, top =textGrob("Cumulative Attack Rate", x = unit(0.5, "npc"), 
                               y = unit(0.7, "npc"))),
                   arrangeGrob(p23, top=textGrob("Maximum Attack Rate", x = unit(0.5, "npc"), 
                               y = unit(.7, "npc"))), 
                   arrangeGrob(p21, top = textGrob("Case Fatality Ratio (CFR)", x = unit(0.5, "npc"), 
                               y = unit(.7, "npc"))), 
                   arrangeGrob(p34, top =textGrob("Relative Start Delay", x = unit(0.5, "npc"), 
                               y = unit(0.7, "npc"))),
                   arrangeGrob(p25, top=textGrob("Initial Growth Period", x = unit(0.5, "npc"), 
                               y = unit(.7, "npc"))),
                   layout_matrix = lay))
```

## B-H correction 

```{r}
pvalues_best<-c(p_best_cfr, p_best_attackrate, p_best_monthlyattackrate, p_best_startdelay, p_best_daystofirst50)
pvalues_best

pvalues_best_BH<-p.adjust(pvalues_best, method="BH")
pvalues_best_BH

pvalues_best_list<-as.data.frame(pvalues_best)
pvalues_best_list

names(pvalues_best_list)<-c("pvalues_best")
names(pvalues_best_list)

pvalues_best_list$pvalues_best_BH<-pvalues_best_BH
pvalues_best_list$pvalues_best_BH
pvalues_best_list

write.csv(pvalues_best_list,"pvalues_best_corrected.csv")

```


# 4. Pairwise correlations between the outcome variables

```{r} 

df_scores <- as.data.frame(res_pca$ind$coord)
df_scores$'Cumulative Attack Rate' = log10(df$cases_per_million)
df_scores$'Maximum Attack Rate' = log10(df$max_monthly_attackrate)
df_scores$CFR = asin(sqrt(df$CFR))
hist(df_scores$CFR)
df_scores$'Relative Start Delay' = sqrt(df$start_delay) 
hist(df_scores$'Relative Start Delay')
df_scores$'Initial Growth Period' = log10(df$days_to_50cases) 
hist(df_scores$'Initial Growth Period')
df_scores$country = df$country

ggpairs(df_scores[, c(5:9)], dpi=300)  
```


